---
title: "Week 3: Discussion Section"
subtitle: "Practice vector operations"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
---

:::{.callout-tip icon=true}
## Reference

Refer to [Chapter 4](https://r.geocompx.org/spatial-operations.html) and [Chapter 5](https://r.geocompx.org/geometry-operations) of *Geocomputation with R* to help get you started.
:::

## Learning Objectives

- Explore topological relationships with `sf::st_intersects()`, `sf::st_intersection()`, `sf::st_within()`, etc.
- Explore distance relationships with `sf::st_distance()`, `sf::st_within_distance()`, and `sf::st_buffer()`
- Learn about spatial and distance-based joins
- Practice writing error/warning messages and unit tests to diagnose outputs

## Data

- **Santa Barbara County's City Boundaries** ([Source](https://geodata.library.ucsb.edu/catalog/3853-s3_2003_s3_cities_polys))
- **California Protected Areas Database** ([Source](https://data.cnra.ca.gov/dataset/california-protected-areas-database))
- **iNaturalist Research-grade Observations, 2020-2024** ([via `rinat`](https://github.com/ropensci/rinat))

## Get Started

Let's load the necessary packages:

```{r}
library(here)
library(tidyverse)
library(sf)
library(tmap)
```

And read in the spatial objects:

```{r}
#| echo: false

sb_protected_areas <- read_sf(here::here("course-materials", "data", "week3-discussion", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")
sb_city_boundaries <- read_sf(here::here("course-materials", "data", "week3-discussion", "sb_city_boundaries_2003.shp")) %>%
  st_transform("ESRI:102009")
sb_county_boundary <- read_sf(here::here("course-materials", "data", "week3-discussion", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")
aves <- read_sf(here::here("course-materials", "data", "week3-discussion", "aves_observations_2020_2024.shp")) %>%
  st_transform("ESRI:102009")
```

```{r}
#| echo: false

sb_protected_areas <- read_sf(here::here("data", "week3-discussion", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")
sb_city_boundaries <- read_sf(here::here("data", "week3-discussion", "sb_city_boundaries_2003.shp")) %>%
  st_transform("ESRI:102009")
sb_county_boundary <- read_sf(here::here("data", "week3-discussion", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")
aves <- read_sf(here::here("data", "week3-discussion", "aves_observations_2020_2024.shp")) %>%
  st_transform("ESRI:102009")
```

To meet our learning objectives, your task:

1. Find how many bird observations are within protected areas in Santa Barbara County
  - Show the different outputs from a spatial subset and a spatial join
  - **Bonus Challenge**: Try it out with a 5 km buffer around the protected areas too! 
2. Find the protected areas within 15 km of a city in Santa Barbara County
  - **Hint**: Use `dplyr::filter()` to select a city from  `sb_city_boundaries`
  - Explore the different outputs with `sf::st_intersects()`, `sf::st_intersection()`, and `sf::st_within()`
  - Practice a distance-based join with `sf::st_is_within_distance()`
3. Find the distance between your city of choice and a protected area of your choice
  - **Note**: `sf::st_distance()` finds the distance between the geometries' edges
