---
title: "Week 1: Discussion Section"
subtitle: "Making maps with `tmap` and beyond"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
---

:::{.callout-tip icon=true}
## Reference

Refer to the [Plotting Geospatial Data](./resources/plotting.qmd) resource and [tmap book](https://r-tmap.github.io/tmap-book/) to help get you started.
:::

## Learning Objectives

-   Read in spatial objects
-   Create map with single spatial object
-   Create map with multiple spatial objects
-   Use different types of `tmap` plotting formats (e.g. `tm_polygons()`, `tm_fill()`, `tm_dots()`, etc.)
-   Adjust color palettes
-   Include essential map elements (e.g. scale bar & north arrow or graticules)
-   Create an interactive map
-   **Bonus Challenge**: Reproduce map using `ggplot2` instead of `tmap`

## Get Started

Let's load the necessary packages:

```{r}
#| message: false
#| warning: false

library(here)
library(tidyverse)
library(stars)
library(sf)
library(tmap)
```

And read in the spatial objects for Easter Island (Rapa Nui/Isla de Pascua):

-   `ei_points`: file contains several points on the island
-   `ei_elev`: raster with elevation data
-   `ei_borders`: polygon with the island outline
-   `ei_roads`: lines contains a road network for the island

```{r}
#| eval: false
#| echo: false

ei_points <- sf::read_sf(here::here("course-materials", "data", "week1", "easter_island", "ei_points.gpkg"))
volcanoes <- subset(ei_points, type == "volcano")
ei_elev <- stars::read_stars(here::here("course-materials", "data", "week1", "easter_island", "ei_elev.tif"))
ei_borders <- sf::read_sf(here::here("course-materials", "data", "week1", "easter_island", "ei_border.gpkg"))
ei_roads <- sf::read_sf(here::here("course-materials", "data", "week1", "easter_island", "ei_roads.gpkg"))
```

```{r}
#| eval: false

ei_points <- sf::read_sf(here::here("data", "week1", "easter_island", "ei_points.gpkg"))
volcanoes <- subset(ei_points, type == "volcano")
ei_elev <- stars::read_stars(here::here("data", "week1", "easter_island", "ei_elev.tif"))
ei_borders <- sf::read_sf(here::here("data", "week1", "easter_island", "ei_border.gpkg"))
ei_roads <- sf::read_sf(here::here("data", "week1", "easter_island", "ei_roads.gpkg"))
```

Now, to meet our learning objectives, your task:

1.  Create a map of Easter Island
2.  Create a map of Easter Island and...

-   ...denote the island's borders and continuous elevation
-   ...denote the island's volcanoes and roads
-   ...play with the color palette and essential map elements

3.  Create an interactive map of Easter Island
