---
title: "Week 5: Discussion Section"
subtitle: "Practice raster operations with vectors"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: true
format:
  html:
    toc: true
---

::: {.callout-note icon=true}
# Source Materials
The following materials are modified from *Geocomputation with R* by Robin Lovelace.
:::

## 1. Learning Objectives

- Use `terra` functions `aggregate()` and `resample()` to create a new raster
- Use `terra` functions `as.polygons()` to convert a raster to a vector of polygons

## 2. Get Started

- Create an `.Rproj` as your version controlled project for Week 5
- Create a Quarto document inside your `.Rproj`
- Load all necessary packages and read spatial objects

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(terra)
library(spData)
library(spDataLarge)
```

```{r}
# Load raster data representing grain sizes with the three classes clay, silt and sand
grain <- terra::rast(system.file("raster/grain.tif", package = "spData"))
```

## 3. Your Task

Now, to meet this week's learning objectives, your task:

1. Subset points higher than 3100 meters in `nz_height` (from `spData`) and create a template raster with `rast()`, where the resolution is 3 km x 3 km for the extent of the subset dataset
  - Count numbers of the highest points in each grid cell
  - Find the maximum elevation in each grid cell
2. With the previous raster, complete the following:
  - Aggregate the raster that counts the highest points in New Zealand/Aotearoa
  - Reduce its geographic resolution by half, such that cells are 6 x 6 km
  - Plot the result
  - Resample back to the original resolution of 3 km x 3 km
3. Polygonize `grain` and filter to only keep squares that represent clay
