---
title: "Week 3: Discussion Section"
subtitle: "Practice vector operations"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
---

## 1. Learning Objectives

- Explore topological relationships with `sf` functions: `st_intersects()`, `st_intersection()`, `st_within()`, etc.
- Explore distance relationships with `sf` functions: `st_distance()`, `st_within_distance()`, and `st_buffer()`
- Learn about spatial and distance-based joins
- Practice writing error/warning messages and unit tests to diagnose outputs

## 2. Get Started

- Create an `.Rproj` as your version controlled project for Week 3
- Create a Quarto document inside your `.Rproj`
- Download [this data folder from Google Drive](https://drive.google.com/file/d/14t3eu5NxZYjwmDVbJr9jnFCrNlsAJVpJ/view) and move it inside your `.Rproj`
- Load all necessary packages and read spatial objects

```{r}
library(here)
library(tidyverse)
library(sf)
library(tmap)
```

```{r}
sb_protected_areas <- read_sf(here::here("data", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")

sb_city_boundaries <- read_sf(here::here("data", "sb_city_boundaries_2003.shp")) %>%
  st_transform("ESRI:102009")

sb_county_boundary <- read_sf(here::here("data", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")

aves <- read_sf(here::here("data", "aves_observations_2020_2024.shp")) %>%
  st_transform("ESRI:102009")
```

## 3. Your Task

You will be wokring with the following datasets:

- **Santa Barbara County's City Boundaries** ([Santa Barbara County](https://geodata.library.ucsb.edu/catalog/3853-s3_2003_s3_cities_polys))
- **California Protected Areas Database** ([CPAD](https://data.cnra.ca.gov/dataset/california-protected-areas-database))
- **iNaturalist Research-grade Observations, 2020-2024** ([via `rinat`](https://github.com/ropensci/rinat))

Now, to meet this week's learning objectives, your task:

1. Find how many bird observations are within protected areas in Santa Barbara County
  - Show the different outputs from a spatial subset and a spatial join
  - **Bonus Challenge**: Try it out with a 5 km buffer around the protected areas too! 
2. Find the protected areas within 15 km of a city in Santa Barbara County
  - **Hint**: Use `dplyr::filter()` to select a city from  `sb_city_boundaries`
  - Explore the different outputs with `st_intersects()`, `st_intersection()`, and `st_within()`
  - Practice a distance-based join with `st_is_within_distance()`
3. Find the distance between your city of choice and a protected area of your choice
  - **Note**: `st_distance()` finds the distance between the geometries' edges
