---
title: "Week 10: Discussion Section"
subtitle: "Practice gridding and spatial interpolation"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
---

::: {.callout-note icon=true}
# Source Materials
The following materials are modified from [Spatial Interpolation in R](https://pages.cms.hu-berlin.de/EOL/gcg_quantitative-methods/Lab14_Kriging.html). 
:::

## 1. Background

Spatial interpolation is a method used to estimate values at unobserved locations based on observed data from nearby points. Two of the most popular spatial interpolation methods are Inverse Distance Weighting (IDW) and ordinary/universal Kriging. Today, you will perform an IDW interpolation, which assumes Tobler's first law of geography, which states that everything is related to everything else, but near things are more related to each other.

## 2. Get Started

- Create an `.Rproj` as your version controlled project for Week 10
- Create a Quarto document inside your project
- Download [this data folder from Google Drive]() and move it inside your project
- Load all necessary packages and read spatial objects

::: {.callout-warning}
If you don't have `gstat` installed, make sure to run:
<br>
```{r}
install.packages("gstat")
```
:::

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(terra)
library(stars)
library(gstat)
library(tmap)
library(kableExtra)
```

```{r}
ca_ecoregions <- st_read(here::here("data", "CA_Level3_Ecoregions_EPA", "ca_eco_l3.shp"))

bioclim_dir <- here::here("data", "wc2.1_2.5m")
bioclim <- list.files(bioclim_dir, pattern = glob2rx("*.tif$"), full.names = TRUE)
bioclim_sort <- bioclim[
  # Sort filepaths based on numeric suffix
  order(
  # Extract numeric suffix of filenames and convert to numeric
  as.numeric(gsub(".*_(\\d+)\\.tif$", "\\1", bioclim)))]
bioclim_rast <- rast(bioclim_sort)

variables <- c("annualMeanTemp", "meanDiurnalRange", "isothermality", "tempSeasonality", "maxTempWarmMonth", "maxTempColdMonth", "tempAnnualRange", "meanTempWetQ", "meanTempDryQ", "meanTempWarmQ", "meanTempColdQ", "annualPrecip", "precipWetMonth", "precipDryMonth", "precipSeasonality", "precipWetQ", "precipDryQ", "precipWarmQ", "precipColdQ")
names(bioclim_rast) <- variables
```

## 3. Your Task

Now, to meet this week's learning objectives, your task:

1. Crop and mask `bioclim_rast` with `ca_ecoregions`
2. Use `terra::spatSample` to generate random points in cropped and masked `bioclim_rast`
    - **Hint**: Set `as.points = TRUE` within `spatSample` and convert to `sf` object for Step #3
3. Create a grid with `sf::st_make_grid` and convert grid to `stars` object with `st_as_stars`
4. Use `gstat::idw` to perform an inverse distance weighted interpolation for a variable of interest (annualMeanTemp, meanDiurnalRange, isothermality, etc.)
5. Use `rast` and `mask` on interpolated `stars` object from Step #4
6. Find interpolated mean of the variable of interest for each ecoregion in California using `zonal`
    - **Hint**: Convert `ca_ecoregions` to `rast` object with `rasterize`