---
title: "Week 1: Discussion Section"
subtitle: "Map making with `tmap` and beyond"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
    code_folding: hide
---

::: callout-tip
## Reference

Refer to the [Plotting Geospatial
Data](course-materials/resources/plotting.qmd) resource to get you
started.
:::

## Learning Objectives

-   Read in spatial objects
-   Create map with single spatial object
-   Create map with multiple spatial objects
-   Use different types of plotting formats (e.g. `tm_polygons`,
    `tm_fill`, `tm_dots`, etc.)
-   Adjust color palettes
-   Include essential map elements (e.g. scale bar & north arrow or
    graticules)
-   Create an interactive map
-   **Bonus Challenge**: Reproduce map using `ggplot2` instead of `tmap`

## Get Started

Let's load the necessary packages:

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(stars)
library(sf)
library(tmap)
```

And read in the spatial objects for Easter Island:

-   **ei_points**: file contains several points on the island
-   **ei_elev**: raster with elevation data
-   **ei_borders**: polygon with the island outline
-   **ei_roads**: lines contains a road network for the island

```{r}
#| eval: false
#| echo: false

ei_points <- sf::read_sf(here::here("course-materials", "data", "week1", "easter_island", "ei_points.gpkg"))
volcanoes <- subset(ei_points, type == "volcano")
ei_elev <- stars::read_stars(here::here("course-materials", "data", "week1", "easter_island", "ei_elev.tif"))
ei_borders <- sf::read_sf(here::here("course-materials", "data", "week1", "easter_island", "ei_border.gpkg"))
ei_roads <- sf::read_sf(here::here("course-materials", "data", "week1", "easter_island", "ei_roads.gpkg"))
```

```{r}
#| eval: false

ei_points <- sf::read_sf(here::here("data", "week1", "easter_island", "ei_points.gpkg"))
volcanoes <- subset(ei_points, type == "volcano")
ei_elev <- stars::read_stars(here::here("data", "week1", "easter_island", "ei_elev.tif"))
ei_borders <- sf::read_sf(here::here("data", "week1", "easter_island", "ei_border.gpkg"))
ei_roads <- sf::read_sf(here::here("data", "week1", "easter_island", "ei_roads.gpkg"))
```

Now, to meet our learning objectives, your task:

1.  Create a map of Easter Island
2.  Create a map of Easter Island and...

-   ...denote the island's borders and continuous elevation
-   ...denote the island's volcanoes and roads
-   ...play with the color palette and essential map elements

3.  Create an interactive map of Easter Island

## Answer Key

The following materials are modified from the [tmap
book](https://r-tmap.github.io/tmap-book/).

### Create a Map of Easter Island
```{r}
#| message: false
#| warning: false
#| eval: false

tm_shape(ei_elev) +
  tm_graticules() +
  tm_raster(col.scale = tm_scale_continuous(),
            col.legend = tm_legend("Elevation (m asl)")) +
  tm_scale_continuous("rd_yl_gn") +
  tm_shape(ei_borders) + 
  tm_borders() +
  tm_shape(ei_roads) +
  tm_lines() +
  tm_shape(volcanoes) +
  tm_symbols(shape = 24, size = "elevation", 
             size.legend = tm_legend("Volcanoes (m asl)")) +
  tm_compass(position = c("right", "top")) +
  tm_scalebar() +
  tm_title("Easter Island")
```

### Create an Interactive Map of Easter Island
```{r}
#| message: false
#| warning: false
#| eval: false

my_map <- tm_shape(ei_elev) +
  tm_raster(col.scale = tm_scale_continuous(),
            col.legend = tm_legend("Elevation (m asl)")) +
  tm_scale_continuous("rd_yl_gn") +
  tm_shape(ei_borders) + 
  tm_borders() +
  tm_shape(ei_roads) +
  tm_lines() +
  tm_shape(volcanoes) +
  tm_symbols(shape = 24, size = "elevation", 
             size.legend = tm_legend("Volcanoes (m asl)")) +
  tm_title("Easter Island")

tmap_mode("view")
my_map
```

