---
title: "Week 5: Discussion Section"
subtitle: "Practice raster operations with vectors"
author: "Alessandra Vidal Meza"
date: last-modified
execute: 
  eval: true
format:
  html:
    toc: true
---

:::{.callout-tip icon=true}
## Reference

Refer to [Chapter 6](https://r.geocompx.org/raster-vector) of *Geocomputation with R* to help get you started.
:::

## Learning Objectives

- Use `terra::crop()` and `terra::mask()` to create a subset of a raster or vector 
- Use `terra::aggregate()` and `terra::resample()` to create a new raster or vector
- Use `terra::as.polygons()` to convert a raster to a vector of polygons

## Get Started

Let's load the necessary packages:

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(terra)
library(spData)
library(spDataLarge)
library(tidyverse)
```

```{r}
# load 30 random points from Zion National Park
zion_points <- read_sf(system.file("vector/zion_points.gpkg", package = "spDataLarge"))

# create a convex hull from the points 
zion_convex_hull <- sf::st_combine(zion_points) %>%
  sf::st_convex_hull() %>% 
  st_as_sf()

# load elevation raster data for Zion National Park, from the Shuttle Radar Topography Mission
srtm <- terra::rast(system.file("raster/srtm.tif", package = "spDataLarge"))

# load raster data representing grain sizes with the three classes clay, silt and sand
grain <- terra::rast(system.file("raster/grain.tif", package = "spData"))
```

Now, to meet our learning objectives, your task:

1. Crop and mask the `srtm` raster using (1) `zion_points` and (2) `zion_convex_hull`
  - Are there any differences in the output maps with `crop()`? What about with `mask()`? 
  - What happens when you set `inverse = TRUE` in `mask()`?
2. Subset points higher than 3100 meters in `nz_height` and create a template raster with `terra::rast()`, where the resolution is 3 km x 3 km for the extent of the subset dataset
  - Count numbers of the highest points in each grid cell
  - Find the maximum elevation in each grid cell
3. With the previous raster, complete the following:
  - Aggregate the raster that counts the highest points in New Zealand/Aotearoa
  - Reduce its geographic resolution by half, such that cells are 6 x 6 km
  - Plot the result
  - Resample back to the original resolution of 3 km x 3 km
4. Polygonize `grain` and filter to only keep squares that represent clay
