---
title: "Week 8: Discussion Section"
subtitle: "Lorem ipsum"
author: "Allie Caughmam"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
---

## Get Started

Let's load the necessary packages:

```{r}
#| message: false
#| warning: false

library(sf)
library(tidyverse)
library(terra)
```

```{r}
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"))
```

And create a raster:

```{r}
elev <- terra::rast(nrows = 6, 
                    ncols = 6, 
                    resolution = 0.5,
                    xmin = -1.5, xmax = 1.5, 
                    ymin = -1.5, ymax = 1.5,
                    vals = 1:36)
```

Now, to meet our learning objectives, your task:

1. Practice filtering a raster
  - Set all cells to NA where `elev < 20`
2. Practice combining (or unioning) geometries
  - Combine all geometries without resolving borders of `nc`
  - Find union of all geometries
  - Remove geometries
3. Create a new geometry that is the difference of two geometries
  - **Hint**: Plot the difference! Counties should be missing!

## Answer Key

### Filter a Raster
```{r}
elev[elev < 20] <- NA
```

### Combine Geometries
```{r}
nc_combine <- sf::st_combine(nc)
plot(nc_combine)

nc_union <- sf::st_union(nc)
plot(nc_union)

counties <- nc %>%
  dplyr::filter(NAME %in% c("Ashe", "Alleghany", "Surry")) %>%
  sf::st_union()

# Plot counties on top of NC unioned
ggplot() +
  geom_sf(data = nc_union, fill = "grey", color = "transparent") +
  geom_sf(data = counties, fill = "black", color = "transparent")
```

### Create a New Geometry
```{r}
nc_difference <- sf::st_difference(nc_union, counties)
nc_disjoint <- sf::st_disjoint(nc_union, counties)
nc_intersection <- sf::st_intersection(nc_union, counties)

ggplot() + 
  geom_sf(data = nc_difference, fill = "grey", color = "transparent") +
  geom_sf(data = nc_intersection, fill = "grey", color = "transparent")
```