---
title: "Discssuion Week 1: Making Maps"
subtitle: "EDS 223: Geospatial Analysis & Remote Sensing"
author: "FirstName LastName"
date: last-modified
---

The following exercises are modified from [Chapter 9 of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/adv-map.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prerequisites

Let's load all necessary packages:

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(terra)
library(spData)
library(spDataLarge)
library(tidyverse)
library(ggspatial)
library(patchwork)
```

## Exercise 1
These exercises rely on a new data object based on the **world** and **worldbank_df** datasets from the **spData** package.

```{r}

```

### Map 1
Create a map showing the geographic distribution of the Human Development Index (HDI) across Africa.  

```{r}

```

```{r}

```

### Map 2
Now, let's update the map so the legend has three bins:
- **High** where HDI is above 0.7
- **Medium** where HDI is between 0.55 and 0.7
- **Low** where HDI is below 0.55

```{r}

```

```{r}

```

### Map 3
Next, let's represent Africa's sub regions on the map:
- Change the color palette 
- Change the legend title 

```{r}

```

```{r}

```

Let's use **patchwork** to combine this map with the previous map into a single plot:

```{r}

```


## Exercise 2
Here, we will use the **zion** and **nlcd** datasets from the **spDataLarge** package:

```{r}

```

```{r eval=FALSE}
# For spDataLarge version 2.0.X
data(nlcd, package = "spDataLarge")
force(nlcd)
```

```{r}
# For spDataLarge version 2.1.1
nlcd <- terra::rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
```

First, we need to convert the raster into a dataframe:
```{r}

```

### Map 4
Let's create a land cover map of Zion National Park:

- Change the default colors to match your perception of land cover categories   
- Move the map legend outside of the map to improve readability 
- Add a scale bar and north arrow
- Change the position of the scale bar and north arrow to improve the maps aesthetics 
- Add the park boundaries on top of the land cover map    

```{r}

```

```{r}

```

### Map 5
Now, let's add an inset of Zion's location in the context of the state of Utah. Hint: an object representing Utah can be subset from the **us_states** dataset.

```{r}

```

```{r}

```

```{r}

```

Let's use **patchwork** again to combine elements:

```{r}

```
