---
title: "Discssuion Week 3: Assignment 2"
subtitle: "EDS 223: Geospatial Analysis & Remote Sensing"
author: "FirstName LastName"
date: last-modified
---

The following exercises are modified from [Chapters 3](https://geocompr.robinlovelace.net/attr.html), [4](https://geocompr.robinlovelace.net/spatial-operations.html), [5](https://geocompr.robinlovelace.net/geometry-operations.html) of Geocomputation with R by Rovin Lovelace.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prerequisites

Let's load all necessary packages:

```{r load, message=FALSE, warning=FALSE}
library(sf)
library(spData)
library(tidyverse)
```

Let's load all necessary data frames:

```{r}
us_states <- us_states %>% 
  janitor::clean_names()

us_states_df <- us_states_df %>% 
  janitor::clean_names()

nz_height <- nz_height %>%  
  janitor::clean_names()

nz <- nz %>%  
  janitor::clean_names()
```

## United States

Let's find the states that belong to the West region:

```{r}

```

And from these states, the states with more than 5,000,000 residents in 2015:

```{r}

```

Next, what was the total population of the US in 2015? 

```{r}

```

Let's now create a new variable named `us_states_stats` by adding variables from **us_states_df** to **us_states**:

```{r}

```

```{r}

```

What is the class of the new object? 

- "sf"
- "data.frame"

Now, calculate the percent change in population density between 2010 and 2015 in each state:

```{r}

```

In how many states did population density decrease? 

```{r}

```

Create a new object representing all of the states the geographically intersect with Colorado.

- Hint: use the **us_states** dataset. The most concise way to do this is with the subsetting method "[".

```{r}

```

Create another object representing all the objects that touch (have a shared boundary with) Colorado and plot the result.

- Hint: remember you can use the argument *op = st_intersects* and other spatial relations during spatial subsetting operations in base R.

```{r}

```

```{r}

```

## New Zealand

How many of New Zealand's high points are in the Canterbury region?

```{r}

```

This could also be achieved by spatial joining then filter

```{r}

```

Which region has the second highest number of **nz_height** points? And how many does it have? 

```{r}

```

Generate simplified versions of the **nz** dataset. Experiment with different values of keep (ranging from 0.5 to 0.00005) for `ms_simplify()` and dTolerance (from 100 to 100,000) for `st_simplify()`.

```{r}

```

Map the results to show how the simplification changes as you change values.

```{r}

```

How many points from the **nz_height** dataset are within 100km of the Canterbury region?

```{r}

```

Find the geographic centroid of the country of New Zealand. How far is it from the geographic centroid of Canterbury?

```{r}

```
