---
title: "Unprofessional Output Example"
author: Juliet Cohen
date: last-modified
execute: 
  eval: true
---

# Unprofessional Documents

Unprofessional documents are messy and leave the reader struggling to follow the story of the analysis.

Examples of components:

- loading packages throughout document
- missing comments, documentation between analysis steps, and data citations
- unnecessary outputs
- intermediate sanity checks retained in final version
- messy code indentation
  - for lists, parameters, functions within functions, etc.
- messy plots, lacking explanation 

# Elephant tracking in Krugar National Park

```{r include = FALSE}
library(tidyverse)

library(here)
library(sf)
```

```{r}
fp <- list.files(path = here("course-materials"), pattern = "elephants.csv", recursive = TRUE, full.names = TRUE)

elephants <- read_csv(fp) %>% 
    sf::st_as_sf(coords =            c("location-long", "location-lat"),
                                  crs = st_crs(4326)) %>%
          filter(st_is_valid(.))

elephants
```

```{r}
elephants_ref <- read_csv(list.files(path = here("course-materials"),
                                     pattern = "elephants_ref_data.csv",
                                     recursive = TRUE,
                                     full.names = TRUE))
```

```{r}
sort(unique(elephants$"individual-local-identifier")) == sort(unique(elephants_ref$"animal-id"))
```

```{r}
library(tmap)
library(rnaturalearth)
library(osmdata)
```

```{r}
south_africa <- rnaturalearth::ne_countries(scale = "medium",
                                                returnclass = "sf") %>%
                                    filter(admin == "South Africa")

tm_shape(south_africa) +
 tm_borders(col = "black", lwd = 0.5) +
 tm_fill(col = "white") +
 tm_shape(elephants) +
tm_dots(col = "individual-local-identifier",
         palette = 'viridis',
         size = 0.1,
         border.col = "black",
         title = "Individual") +
 tm_layout(
   bg.color = "lightblue",
   title = "Elephant Observations",
   frame = TRUE,
   title.position = c(0.01, 0.5),
   title.size = 1.2,
   legend.frame = TRUE,
   legend.position = c(0.01, 0.2)
 )
```

All of the elephants were in South Africa.

Data from somewhere on MoveBank: https://datarepository.movebank.org

